**Evaluation of machine learning methods for 16S rRNA gene data**

### 2019-06-13 submission/manuscript.Rmd
discussion additions

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce in part due to the flawed modeling methods that call into question the validity of developed ML models. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

As the complexity of a ML model increase, its training times will increase as well (Table S1). When regularization and hyperparameter optimization (i.e. model selection) steps are added to the model pipeline, training times will increase further. The more hyperparameters a model has, the longer it will take to train. The computational burden that comes from model complexity can be reduced by parallelizing different segments of the ML pipeline. In this study we have parallelized each data-split which allowed 100 data-splits to be run at the same time for each model. We can further parallelize the cross-validation step for each hyper-parameter setting.

@@

- In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the differences between cross-validation and test performance is low, this suggets the models will perform similar with similar data. In all seven models, the differences in cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models will be able to test similar new data.

- We also reported the results of 100 random dataset splits because there are differences in AUROC values of each of the seven models between different random data-splits. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split. This result shows that you can get "lucky" in terms of prediction performance depending on the random split. Therefore, it is crucial to report all the findings over many random data-splits to get reliable performance metrics.

 - Our results also showed that we can choose different models for different reasons. Getting feature importances as weights and regression coefficients is the fastest and easiest interpratation. Similarly, linear models require less computational burden, fastest to train. Thinking about the applicability, and actionability of a model in clinical setting, we might want faster results.

 - There is variability if we don't do a full grid search for the hyper-parameters. We can give each model 15 minutes to train and see how well they do. So an L1 linear model will give you terible results with one and not the other C value as seen in figure S1. We have done a granular search first and then limited the possible settings to a range that made the more sense.

 - Suprisingly, l2 logistic regression does really well.

 - Computational efficiency and interp of logistic vs random forest

 - decision tree and l1 why so low?  Decision tree model has no regularization which is why the AUROC values were low.

>>>>

Machine learning (ML) modeling of the human microbiome has the potential to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but rigourous ML models are scarce due to the flawed modeling methods that call into question the validity of developed ML models. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained much faster (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

As the complexity of a ML model increase, its training times will increase as well (Table S1). When regularization and hyperparameter optimization (i.e. model selection) steps are added to the model pipeline, training times will increase further. The more hyperparameters a model has, the longer it will take to train. The computational burden that comes from model complexity can be reduced by parallelizing different segments of the ML pipeline. In this study we have parallelized each data-split which allowed 100 data-splits to be processed through the ML pipeline at the same time for each model. We can further parallelize the cross-validation step for each hyper-parameter setting.

@@

In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the differences between cross-validation and test performance is low, this suggets the models will perform similar with similar data. In all seven models, the differences in cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models will be able to test similar new data.

In our ML pipeline we performed the 80%-20% random datasplit 100 times. The repetition of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop generalizable ML models and to report reliable performance metrics for a ML model. Depending on the data-split, there is the chance of being overoptimistic about the predictive performance of a model. In our study, we showed that there are differences in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split.

Our results also showed that we should choose to use ML models based on the goal of the study and our expectations of predictive performance, interpretibility and computational burden. In terms of predictive performance, all models except a decision tree and L1-regularized SVM with linear kernel behaved similarly. Random forest model had testing AUROC values statistically significantly larger than others. However the second best model was L2-regularized logistic regression following random forest with a median AUROC difference of only 0.015. The decision tree model performed poorly because it had no regularization. L1-regularized SVM with linear kernel also performed poorly possibly because it was more sensitive to hyper-parameter tuning. In terms of interpretation, random forest is a more complex ML model and it can only be explained using methods such as permutation importance. On the other hand, L2-regularized logistic regression is easy to interpret (i.e. regression coefficients of the trained model). If the goal of the study is to learn the ecology behind a disease and to identify biomarkers, then easily interpretable models could be a better fit. It is also important to consider the computational burden of developing ML models. Random forest model was slow to train with upto `r` hours whereas L2-regularized logistic regression trained in `r` minutes.

Our study underscored the importance of performing a full grid search for hyperparameter settings when building a ML model. We first performed a granular grid search and then a full grid search on the limited range of possible hyperparameter settings. In the example of L1-regularized SVM with linear kernel, the model showed variability between different regularization coefficients (C).
